---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

An R package for maintaining a local collection of sparkly data sets.

See [RAADTools](https://github.com/AustralianAntarcticDivision/raadtools) for reading, plotting, and manipulating these data.

## Installing

```{r,eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("AustralianAntarcticDivision/bowerbird")
```

## Usage

`bowerbird` must be configured to tell it which data sets to synchronise and where to save them on the local file system.


Build up a configuration by choosing from data sources that are provided as part of the package:

```{r,eval=FALSE}
cf <- bb_config(local_file_root="/your/data/directory",
                wget_flags="--no-passive-ftp") %>%
    add(bb_sources() %>% dplyr::filter(id=="seaice_smmr_ssmi_nasateam"))
```

Or define your own data source:
```{r,eval=FALSE}
my_source <- bb_source(
    name="GSHHG coastline data",
    description="A Global Self-consistent, Hierarchical, High-resolution Geography Database",
    reference= "http://www.soest.hawaii.edu/pwessel/gshhg",
    citation="Wessel, P., and W. H. F. Smith, A Global Self-consistent, Hierarchical, High-resolution Shoreline Database, J. Geophys. Res., 101, 8741-8743, 1996",
    source_urls="ftp://ftp.soest.hawaii.edu/gshhg/*",
    license="",
    comment="",
    method="wget",
    method_flags="--recursive --level=1 --accept=\"*bin*.zip,README.TXT\"",
    postprocess="unzip")

cf <- bb_config(local_file_root="/your/data/directory") %>%
    add(my_source)
```


Load and save configurations directly:

```{r,eval=FALSE}
cf <- ... ## set up configuration to your liking

## save it for direct use later
bb_save_config(cf,"/path/to/your/config.json")

## load up a saved configuration
cf <- bb_config(file="/path/to/your/config.json")
```

And once the configuration is defined, run the sync process:
```{r,eval=FALSE}
bb_sync(cf)
```
